Computing the bicharacter of huge subspaces of polynomials

### 2019-07-24 article/article.tex
Complete section on diagonal hamonics with inert variables.

% Maths
\newcommand{\RR}{\mathbb{R}}
\newcommand{\KK}{\mathbb{K}}
\newcommand{\NN}{\mathbb{N}}
\newcommand{\ZZ}{\mathbb{Z}}
\newcommand{\QQ}{\mathbb{Q}}
\newcommand{\CC}{\mathbb{C}}
\newcommand{\Sym}[1]{\mathbb{S}_{#1}}

@@

\nicolas{Ajouter des exemples au fur et à mesure}

@@

	Throughout this article, we will write the spaces and their the character the same way, except if there might be a confusion, then the variables will be added to the character and we will write $\mathcal{H}_n^{(r)}(\bm{q};\bm{w})$ for the character of $\mathcal{H}_n^{(r)}$,  with $\bm{q}$ the variables used to express the $\GLr$ part of the character and $\bm{w}$ for the $\Sn$ part.

	In the next section, we expose strategies used to go further in the computation of those characters. In the last section, we present some applications/generalizations we are able to make using the same computation strategies. \pauline{annonce du contenu à réécrire}

@@

	In fact we can do better. \pauline{Does it work only for $\harmonics_n^{(r)}$ or also for $\mathcal{E}_\lambda$?}
	Fix a partition $\mu$. Recall that a basis of the Specht module $S_\mu$  is given by the family of Garnir/Specht polynomials $(G_{P})_P$ indexed by standard tableaux $P$ of shape $\mu$.

@@

	\section{Applications}

@@

	\begin{figure}[H]
		\centering
		\begin{tikzpicture}
		\draw node{$\begin{pmatrix}
			x_{1,1} & x_{1,2} & \cdots & x_{1,n} \\
			x_{2,1} & x_{2,2} & \cdots & x_{2,n} \\
			\vdots & \vdots & \ddots & \vdots \\
			x_{r,1} & x_{r,2} & \cdots & x_{r,n} \\ 
			\theta_{1} & \theta_{2} & \cdots & \theta_{n}
			\end{pmatrix}$ } (0,0);
		\draw (-2.4,.6)+(60:.25)
		[xscale=.5,yscale=3.5,->,rotate=80] arc(-40:290:.25);
		\node at (-3,0.3) {{\small$\GLr$}};
		\draw (-.6,-1.7)+(60:.25)
		[xscale=5.8,yscale=.6,->,rotate=-210] arc(-40:290:.25);
		\node at (0,-2.1) {{\small$\Sn$}};
		\draw (-3,-.85) -- (-1.5,-.85);
		\end{tikzpicture}
		\caption{Simultaneous actions of $\GLr$ and $\Sn$ on $X$ with inert variables.}
	\end{figure}

@@

	Like before, we consider the smallest subspace containing $V_{\lambda}$ and closed by partial derivation and polarization, except that this time the inert variables do not intervene in these operators. We denote this space by $\mathcal{E}_{\lambda}^{(r)}$. The space of diagonal harmonics $\harmonics_n^{(r)}$ is thus a special case of $\mathcal{E}_{\lambda}^{(r)}$ for $\lambda = (n)$. \\

	The character of $\mathcal{E}_{\lambda}^{(r)}$ stabilizes for $r \geq n$ (see \cite{Bergeron2013} or \cite{Bergeron2009}). Therefore, when the precision is not necessary, we will drop the exponent and simply write $\mathcal{E}_{\lambda}$ for $\mathcal{E}_{\lambda}^{(r)}$ with $r \geq n$. \\

	Like for $\harmonics_n^{(r)}$, the $\GLr \times \Sn$-character of $\mathcal{E}_\lambda^{(r)}$ is not easy to compute but hopefully most of the strategies we used to compute the character of $\harmonics_{n}^{(r)}$ work for $\mathcal{E}_\lambda^{(r)}$.

	Therefore, we, once again, use polarization to compute $\mathcal{E}_\lambda^{(r)}$ from $\mathcal{E}_\lambda^{(1)}$, exploit the multigrading. We use the same isotypic projector to obtain the isotypic components, then we can still exploit the columns antisymmetries. 
%	\begin{example} ~
%		
%		$V_{21} = -x_{2} \theta_{1} + x_{3} \theta_{1} + x_{1} \theta_{2} - x_{3} \theta_{2} - x_{1} \theta_{3} + x_{2} \theta_{3}$ is of type $(1,1,1)$ and has reduced form $x_1\theta_2$.
%		
%		$\partial_{x_2}V_{21} = -\theta_1 + \theta_3$ is of type $(2,1)$ and has reduced form $-\theta_1$.
%	\end{example}
	Finally, $\mathcal{E}_\lambda^{(r)}$ is still symmetric with respect to a permutation of the rows of $X$ containing classical variables so we exploit row symmetries in this case too.\\

	\begin{example} In the following example, we use different strategies to compute $\mathcal{E}_{2,1}^{(2)}$.

		First, we compute $\mathcal{E}_{2,1}^{(1)}$, for one row of variables exploiting multidegree. 
		\begin{sagecommandline}
			sage: v = vandermonde(Partition([2,1]))
			sage: v
			sage: derivatives = partial_derivatives(v.parent())
			sage: derivatives
			sage: E_21 = Subspace({v.multidegree(): [v]}, derivatives)
			sage: E_21.basis()
			sage: E_21.dimension()
		\end{sagecommandline}
		And we project on isotypic component, using antisymmetries. 
		\begin{sagecommandline}
			sage: E_21 = IsotypicComponent(E_21, 3,use_antisymmetry=True)
			sage: E_21
		\end{sagecommandline}
		Then, we use polarization operators to introduce a second set of variables, and thus obtain $\mathcal{E}_{2,1}^{(2)}$, still with multidegree. We also exploit row symmetries. 
		\begin{sagecommandline}
			sage: polarizators = polarization_operators(2, row_symmetry="permutation")
			sage: E_21 = PolarizedSpace(E_21, polarizators)
			sage: E_21[Partition([1,1,1])].basis()
			sage: E_21[Partition([2,1])].basis()
		\end{sagecommandline}
		And we easily obtain the character for $\mathcal{E}_{2,1}$. 
	 	\begin{sagecommandline}
		 	sage: character(E_21, row_symmetry="permutation")
		\end{sagecommandline}

	So the character of $\mathcal{E}_{2,1}^{(r)}$ for $r \geq 2$ is $$\mathcal{E}_{2,1}^{(r)} = 1 \otimes s_{2,1} + s_{1} \otimes s_{1,1,1} .$$
	\end{example}

	\subsection{Further work}

>>>>

% Maths
\newcommand{\RR}{\mathbb{R}}
\newcommand{\KK}{\mathbb{K}}
\newcommand{\NN}{\mathbb{N}}
\newcommand{\ZZ}{\mathbb{Z}}
\newcommand{\QQ}{\mathbb{Q}}
\newcommand{\CC}{\mathbb{C}}
\newcommand{\Sym}[1]{\mathbb{S}_{#1}}
\newcommand{\Elambda}[1]{\mathcal{E}_{#1}}
\newcommand{\Elambdar}[2]{\mathcal{E}_{#1}^{(#2)}}

@@

\nicolas{Ajouter des exemples au fur et à mesure}
\pauline{J'ai déjà mis pas mal d'exemples. Il en faut plus?}

@@

	Throughout this article, we will write the spaces and their the character the same way, except if there might be a confusion, then the variables will be added to the character and we will write $\mathcal{H}_n^{(r)}(\bm{q};\bm{w})$ for the character of $\mathcal{H}_n^{(r)}$,  with $\bm{q}$ the variables used to express the $\GLr$ part of the character and $\bm{w}$ for the $\Sn$ part.\\

	In the next section, we expose strategies used to go further in the computation of those characters. Those strategies are not limited to the diagonal harmonic polynomials. They can be used in other generalizations. We introduced some of these generalizations in the last section and discuss some applications of these computations.

@@

	In fact we can do better. \pauline{Does it work only for $\harmonics_n^{(r)}$ or also for $\Elambda{\lambda}$?}
	Fix a partition $\mu$. Recall that a basis of the Specht module $S_\mu$  is given by the family of Garnir/Specht polynomials $(G_{P})_P$ indexed by standard tableaux $P$ of shape $\mu$.

@@

	\section{Applications / Generalizations}

@@

	Note that inert variables, like constants, do not contribute to the degree of the polynomials in which they intervene. They can be considered as "degree 0 variables".

	\begin{figure}[H]
		\centering
		\begin{tikzpicture}
		\draw node{$\begin{pmatrix}
			x_{1,1} & x_{1,2} & \cdots & x_{1,n} \\
			x_{2,1} & x_{2,2} & \cdots & x_{2,n} \\
			\vdots & \vdots & \ddots & \vdots \\
			x_{r,1} & x_{r,2} & \cdots & x_{r,n} \\ 
			\theta_{1} & \theta_{2} & \cdots & \theta_{n}
			\end{pmatrix}$ } (0,0);
		\draw (-2.4,.6)+(60:.25)
		[xscale=.5,yscale=3.5,->,rotate=80] arc(-40:290:.25);
		\node at (-3,0.3) {{\small$\GLr$}};
		\draw (-.6,-1.7)+(60:.25)
		[xscale=5.8,yscale=.6,->,rotate=-210] arc(-40:290:.25);
		\node at (0,-2.1) {{\small$\Sn$}};
		\draw (-1.8,-.85) -- (1.8,-.85);
		\end{tikzpicture}
		\caption{Simultaneous actions of $\GLr$ and $\Sn$ on classical and inert variables.}
	\end{figure}

@@

	Like before, we consider the smallest subspace containing $V_{\lambda}$ and closed by partial derivation and polarization, except that this time the inert variables do not intervene in these operators. We denote this space by $\Elambdar{\lambda}{r}$. The space of diagonal harmonics $\harmonics_n^{(r)}$ is thus a special case of $\Elambdar{\lambda}{r}$ for $\lambda = (n)$. \\

	The character of $\Elambdar{\lambda}{r}$ stabilizes for $r \geq n$ (see \cite{Bergeron2013} or \cite{Bergeron2009}). Therefore, when the precision is not necessary, we will drop the exponent and simply write $\Elambda{\lambda}$ for $\Elambdar{\lambda}{r}$ with $r \geq n$.

	This property also gives use an upper bound for the number of sets of variables needed in the computation. \\

	Like for $\harmonics_n^{(r)}$, the $\GLr \times \Sn$-character of $\Elambdar{\lambda}{r}$ is not easy to compute due to the dimension but hopefully most of the strategies we used to compute the character of $\harmonics_{n}^{(r)}$ work for $\Elambdar{\lambda}{r}$.

	Therefore, we, once again, use polarization to compute $\Elambdar{\lambda}{r}$ from $\Elambda{\lambda}{1}$ and exploit the multigrading. We use the same isotypic projector to obtain the isotypic components, then we can still exploit the columns antisymmetries. 
%	\begin{example} ~
%		
%		$V_{21} = -x_{2} \theta_{1} + x_{3} \theta_{1} + x_{1} \theta_{2} - x_{3} \theta_{2} - x_{1} \theta_{3} + x_{2} \theta_{3}$ is of type $(1,1,1)$ and has reduced form $x_1\theta_2$.
%		
%		$\partial_{x_2}V_{21} = -\theta_1 + \theta_3$ is of type $(2,1)$ and has reduced form $-\theta_1$.
%	\end{example}
	Finally, $\Elambdar{\lambda}{r}$ is still symmetric with respect to a permutation of the rows of $X$ containing classical variables so we exploit row symmetries in this case too.\\

	\begin{example} In the following example, we use different strategies to compute $\Elambdar{2,1}{2}$.

		First, we compute $\Elambdar{2,1}{1}$, for one row of variables exploiting multidegree. 
		\begin{sagecommandline}
			sage: v = vandermonde(Partition([2,1]))
			sage: v
			sage: derivatives = partial_derivatives(v.parent())
			sage: derivatives
			sage: E_21 = Subspace({v.multidegree(): [v]}, derivatives)
			sage: E_21.basis()
			sage: E_21.dimension()
		\end{sagecommandline}
		And we project on isotypic component, using antisymmetries. 
		\begin{sagecommandline}
			sage: E_21 = IsotypicComponent(E_21, 3,use_antisymmetry=True)
			sage: E_21
		\end{sagecommandline}
		Then, we use polarization operators to introduce a second set of variables, and thus obtain $\Elambdar{2,1}{2}$, still with multidegree. We also exploit row symmetries. 
		\begin{sagecommandline}
			sage: polarizators = polarization_operators(2, row_symmetry="permutation")
			sage: E_21 = PolarizedSpace(E_21, polarizators)
			sage: E_21[Partition([1,1,1])].basis()
			sage: E_21[Partition([2,1])].basis()
		\end{sagecommandline}
		And we obtain the character for $\Elambda{2,1}$. 
	 	\begin{sagecommandline}
		 	sage: character(E_21, row_symmetry="permutation")
		\end{sagecommandline}

	So the character of $\Elambdar{2,1}{r}$ for $r \geq 2$ is $$\Elambdar{2,1}{r} = 1 \otimes s_{2,1} + s_{1} \otimes s_{1,1,1} .$$
	\end{example}

	~

	We presented in the previous examples all the steps of the computation using or not different strategies to optimize the computations. Thus, the code can be easily adapted to other generators and other operators (see section \ref{section_futher_work} for examples) at each step.

	Therefore, to simply compute the character of $\Elambda{\mu}$ only setting the optimizations used, one can use the function \texttt{E\_mu} that takes in argument a partition or a diagram and returns the complete character $\Elambda{\mu}$. By default, this function uses antisymmetries of column and symmetries of rows, but this can be disabled.

	Furthermore, parallel computation can be activate in this function. Indeed, once $\Elambda{\lambda}{1}$ is projected on its isotypic component, the computation is independent on each isotypic component and the character is simply the sum of the characters of all isotypic components.

	\begin{example} \pauline{Ajouter d'autres exemples pertinents.}
		We compute $\Elambda{1,1,1}$

		\begin{sagecommandline}
			sage: E_mu(Partition([1,1,1]))
		\end{sagecommandline}	
	\end{example}

	\subsection{Further work}\label{section_futher_work}
